### 2015-03-27 Kokkos_PG_Views.tex
Worked on view section
this cleans up and reorders some sections to make the flow clearer.
I also expanded a few subsections and corrected errenous statements.

@@

\subsection{Memory management and copying}

\subsubsection{Reference counting}

@@

\subsubsection{Deep copy}

If you want to make a deep copy of a View's contents, you must do so
explicitly, using the \lstinline!deep_copy! function.  For example:
\begin{lstlisting}
View<int*> a ("a", 10);
View<int*> b ("b", 10);
deep_copy (a, b); // copy contents of b into a
\end{lstlisting}

\subsection{Where and how is the View allocated?}

\subsubsection{Execution and memory spaces}

Since there is no bijective association between execution spaces and memory 
spaces, Kokkos provides a way to explicitly provide both to a View as a 
\lstinline|Device|. 
\begin{lstlisting}
View<int*, Device<Cuda,CudaUVMSpace> > a ("a", 100000);
View<int*, Device<OpenMP,CudaUVMSpace> > b ("b", 100000);
\end{lstlisting}
In this cas \lstinline|a| and \lstinline|b| will live in the same memory space,
but \lstinline|a| will be initialized on the GPU while \lstinline|b| will be 
initialized on the host.

\subsubsection{Initialization}

A View's entries are initialized to zero by default.  Initialization
happens in parallel, for first-touch allocation over the first
(leftmost) dimension of the View.

@@

Kokkos selects a View's default layout for optimal parallel access
over the leftmost dimension, based on its execution space.  For
example, \lstinline!View<int**, Cuda>! has \lstinline!LayoutLeft!, so that
consecutive threads in the same warp access consecutive entries in
memory.  This \emph{coalesced access} gives the code better memory
bandwidth.  In contrast, \lstinline!View<int**, OpenMP>! has
\lstinline!LayoutRight!, so that a single thread accesses contiguous
entries of the array.  This avoids wasting cache lines and helps
prevent false sharing of a cache line between threads.

@@

TODO: Hey Christian, what did you mean by this?

Christian says: \lstinline!CudaSpace!, \lstinline!HostSpace!, etc.
Also, please emphasize memory space over execution space, and talk
about the UVM and pinned variants of \lstinline!CudaSpace!.

@@

Different optional template parameters of View let you control how
Kokkos accesses entries of the array.  The simplest example involves
making a View \lstinline!const!, that is, making its entries read-only.  It
is a compile-time error to assign to an entry of a const View.  It
does not make sense to allocate a const View, but it does make sense
to assign a non-const View to a const View.  Here is an example:
\begin{lstlisting}
const size_t N0 = ...;
View<double*> a_nonconst ("a_nonconst", N0);

// Assign a nonconst View to a const View
View<const double*> a_const = a_nonconst;
// Pass the const View to some read-only function.
const double result = readOnlyFunction (a_const);
\end{lstlisting}
Note that a ``const View'' means the \emph{entries} are const.  You
may still assign to a const View.  \lstinline!View<const double*>!
corresponds exactly to \lstinline!const double*!, and
\lstinline!const View<double*>! to \lstinline!double* const!.

Const Views often allow the compiler to optimize more agressively by allowing it
to reason about possible write conflicts and data aliasing. For example in a vector
update \lstinline|a(i+1)+=b(i)| with skewed indexing it is safe to vectorize if
\lstinline|b| is a View of const data.

\subsection{Memory Traits}

Another way to get optimized data accesses is to specify a memory trait. 
These traits are used to declare intended use of the particular view of 
an allocation. For example a particular kernel might use a view only for 
streaming writes. By declaring that intention Kokkos can insert the 
appropriate store intrinsics on each architecture if available.

Currently 
In particular, Kokkos lets you declare that you plan
to access a read-only array irregularly (less spatial locality), using
the optional \lstinline!RandomAccess! template parameter.  You may only use
this with a const View.  For example:
\begin{lstlisting}
const size_t N0 = ...;
View<int*> a_nonconst ("a", N0); // allocate nonconst View
// Assign to const, RandomAccess View
View<const int*, RandomAccess> a_ra = a_nonconst;
\end{lstlisting}
For a View whose execution space is \lstinline!Cuda!, access to a
\lstinline!RandomAccess! View may use CUDA texture fetches.  Texture
fetches are not cache coherent with respect to writes, which is why
you must use read-only access.  The texture cache is optimized for
noncontiguous access, since it has a shorter cache line than the
regular cache.

\lstinline!RandomAccess! is valid for other execution spaces, and Kokkos
also requires a const View for those execution spaces.  Kokkos
reserves the right to perform analogous optimizations on other
execution spaces.

@@

It's always better to let Kokkos control memory allocation, but
sometimes you don't have a choice.  You might have to work with an
application or an interface that returns a raw pointer, for example.
Kokkos lets you wrap raw pointers in an \emph{unmanaged View}.
``Unmanaged'' means that Kokkos does \emph{not} do reference counting
or automatic deallocation for those Views.  The following example
shows how to create an unmanaged View of host memory.  You may do this
for CUDA device memory too, or indeed for memory allocated in any
memory space, by specifying the View's execution or memory space
accordingly.
\begin{lstlisting}
// Sometimes other code gives you a raw pointer, ...
const size_t N0 = ...;
double* x_raw = malloc (N0 * sizeof (double));
{
  // ... but you want to access it with Kokkos.
  //
  // malloc() returns host memory, so we use the 
  // host memory space HostSpace.  Unmanaged
  // Views have no label, because labels work with
  // the reference counting system.
  View<double*, HostSpace, Unmanaged> x_view (x_raw, N0);

@@

The \lstinline!Atomic! access trait lets you create a View, such that every
read or write to any entry uses an atomic update.  This only works for
certain data types, for which the hardware and programming model
support atomic updates.  Currently, Kokkos limits atomic updates to
32-bit or 64-bit built-in types.

>>>>

A view can have const data semantics (i.e. its entries are read-only) by 
specifying a `const` data type.  It is a compile-time error to assign to an 
entry of a ``const View''.  Assignment semantics are equivalent to a pointer to 
const data.  
A const View means the \emph{entries} are const.  You
may still assign to a const View.  \lstinline!View<const double*>!
corresponds exactly to \lstinline!const double*!, and
\lstinline!const View<double*>! to \lstinline!double* const!.
Therefore it does not make sense to allocate a const View,
since you could not obtain a non-const view of the same data and you 
can not assign to it. You can however assign a non-const view to a const
view.  Here is an example:
\begin{lstlisting}
const size_t N0 = ...;
View<double*> a_nonconst ("a_nonconst", N0);

// Assign a nonconst View to a const View
View<const double*> a_const = a_nonconst;
// Pass the const View to some read-only function.
const double result = readOnlyFunction (a_const);
\end{lstlisting}

Const Views often enables the compiler to optimize more aggressively by allowing it
to reason about possible write conflicts and data aliasing. For example in a vector
update \lstinline|a(i+1)+=b(i)| with skewed indexing it is safe to vectorize if
\lstinline|b| is a View of const data.

@@

\subsection{Reference counting}

@@

\subsection{Resizing}

Kokkos Views can be resized using the \lstinline|resize| non-member function.
It takes an existing view as its input by reference and the new dimension information
corresponding to the constructor arguments. A new view with the new dimensions will
be created and a kernel will be run in the views execution space to copy the data 
element by element from the old view to the new one. Note that the old allocation is 
only deleted if the view to be resized was the {\it only} view referencing the underlying 
allocation.

\begin{lstlisting}
// Allocate a view with 100x50x4 elements 
View<int**[4]> a( "a", 100,50);

// Resize a to 200x50x4 elements; the original allocation is freed
resize(a, 200,50);

// Create a second view b viewing the same data as a
View<int**[4]> b = a;
// Resize a again to 300x60x4 elements; b is still 200x50x4
resize(a,300,60);
\end{lstlisting}

@@

Kokkos selects a View's default layout for optimal parallel access
over the leftmost dimension, based on its execution space.  For
example, \lstinline!View<int**, Cuda>! has \lstinline!LayoutLeft!, so that
consecutive threads in the same warp access consecutive entries in
memory.  This \emph{coalesced access} gives the code better memory
bandwidth.  In contrast, \lstinline!View<int**, OpenMP>! has
\lstinline!LayoutRight!, so that a single thread accesses contiguous
entries of the array.  This avoids wasting cache lines and helps
prevent false sharing of a cache line between threads. 
In section \ref{S:View:Placement} more details will be discussed.

@@

\subsection{Memory spaces}

Since there is no bijective association between execution spaces and memory 
spaces, Kokkos provides a way to explicitly provide both to a View as a 
\lstinline|Device|. 
\begin{lstlisting}
View<int*, Device<Cuda,CudaUVMSpace> > a ("a", 100000);
View<int*, Device<OpenMP,CudaUVMSpace> > b ("b", 100000);
\end{lstlisting}
In this case \lstinline|a| and \lstinline|b| will live in the same memory space,
but \lstinline|a| will be initialized on the GPU while \lstinline|b| will be 
initialized on the host. 
The \lstinline|Device| type can be accessed as a views \lstinline|device_type|
typedef. A \lstinline|Device| has only three typedef members 
\lstinline|device_type|, \lstinline|execution_space|, and \lstinline|memory_space|.
The \lstinline|execution_space|, and \lstinline|memory_space| typedefs are
the same for a view and the \lstinline|device_type| typedef.

It is important to understand that accessibility of a View does not 
depend on its execution space directly. It is only determined by its
memory space. Therefore both \lstinline|a| and \lstinline|b| have the 
same access properties. They differ only in how they are initialized, 
as well as in where parallel kernels associated with operations such 
as resizing or deep copies are run.

\subsection{Initialization}

A View's entries are initialized to zero by default.  Initialization
happens in parallel, for first-touch allocation over the first
(leftmost) dimension of the View using the execution space
of the View.

\subsection{Deep copy and HostMirror}

Copying data from one view to another, in particular between views 
in different memory spaces, is called deep copy. 
Kokkos never performs a hidden deep copy. To do so a user has to 
call the \lstinline!deep_copy! function.  For example:
\begin{lstlisting}
View<int*> a ("a", 10);
View<int*> b ("b", 10);
deep_copy (a, b); // copy contents of b into a
\end{lstlisting}

Deep copies can only be performed between views with an identical 
memory layout and padding. For example the following two operations
are not valid:
\begin{lstlisting}
View<int*[3]> a ("a", CudaSpace, 10);
View<int*[3]> b ("b", HostSpace, 10);
deep_copy (a, b); // This will give a compiler error

View<int*[3], LayoutLeft> c ("c", CudaSpace, 10);
View<int*[3], LayoutLeft> d ("d", HostSpace, 10);
deep_copy (c, d); // This might give a runtime error
\end{lstlisting}

The first one will not work because the default layouts of \lstinline|CudaSpace|
and \lstinline|HostSpace| are different. The compiler will catch that since 
no overload of the \lstinline|deep_copy| function exists to copy view from
one layout to another. The second case will fail at runtime if padding settings
are different for the two memory spaces. This would result in different 
allocation sizes and thus prevent a direct memcopy.

The reasoning for allowing only direct bitwise copies is that a deep copy
between different memory spaces would otherwise require a temporary 
copy of the data to which a bitwise copy is performed followed by a parallel
kernel to transfer the data element by element.

Kokkos provides the following way to work around those limitations. 
First views have a \lstinline|HostMirror| typedef which is a view type with 
compatible layout inside the \lstinline|HostSpace|. Additionally there is a
\lstinline|create_mirror| and \lstinline|create_mirror_view| function which 
allocate views of the \lstinline|HostMirror| type of a view. 
The difference between the two is that \lstinline|create_mirror| will always
allocate a new view, while \lstinline|create_mirror_view| will only create a 
new view if the original one is not in \lstinline|HostSpace|.

\begin{lstlisting}
View<int*[3]> a ("a", MemorySpace, 10);
// Allocate a view in HostSpace with the 
// layout and padding of a
typename View<int*[3]>::HostMirror b =
    create_mirror(a);
// This is always a memcopy
deep_copy (a, b);

typename View<int*[3]>::HostMirror c =
    create_mirror_view(a);
// This is a no-op if MemorySpace is HostSpace
deep_copy (a,c)
\end{lstlisting}

@@

Another way to get optimized data accesses is to specify a memory trait. 
These traits are used to declare intended use of the particular view of 
an allocation. For example a particular kernel might use a view only for 
streaming writes. By declaring that intention Kokkos can insert the 
appropriate store intrinsics on each architecture if available. Access traits
are specified through an optional template parameter which comes last in the list
of parameters. Multiple traits can be combined with binary ``or'' operators:

\begin{lstlisting}
View<double*, MemoryTraits<SomeTrait> > a;
View<const double*, MemoryTraits<SomeTrait | SomeOtherTrait> > b;
View<double*, LayoutLeft, MemoryTraits<SomeTrait | SomeOtherTrait> > c;
View<double*, MemorySpace, MemoryTraits<SomeTrait | SomeOtherTrait> > d;
View<double*, LayoutLeft, MemorySpace, MemoryTraits<SomeTrait> > e;
\end{lstlisting}

The \lstinline!Atomic! access trait lets you create a View of data , such that every
read or write to any entry uses an atomic update.  This only works for
certain data types, for which the hardware and programming model
support atomic updates.  Currently, Kokkos limits atomic updates to
32-bit or 64-bit types\footnote{On X86 architectures Kokkos also supports 
128-bit atomics. Types for which atomic access are performed must support the 
necessary operators such as =,+=,-=,+,- etc.}.

Note that it is a violation of the semantics to write \emph{or} read from a 
non-atomic view of data if at the same time an atomic view of the data is used.

\begin{lstlisting}
View<int*> a("a" , 100);
View<int*, MemoryTraits<Atomic> > a_atomic = a;

a_atomic(1) += 1; // This access will do an atomic addition
\end{lstlisting}

\subsection{Random Access}

The \lstinline|RandomAccess| trait declares the intent to access a View
irregularly (in particular non consecutively). If used for a const View 
in the \lstinline|CudaSpace| or \lstinline|CudaUVMSpace|, Kokkos will use 
texture fetches for accesses when executing in the \lstinline|Cuda| execution
space.
For example:
\begin{lstlisting}
const size_t N0 = ...;
View<int*> a_nonconst ("a", N0); // allocate nonconst View
// Assign to const, RandomAccess View
View<const int*, RandomAccess> a_ra = a_nonconst;
\end{lstlisting}
If the default execution space is \lstinline!Cuda!, access to a
\lstinline!RandomAccess! View may use CUDA texture fetches.  Texture
fetches are not cache coherent with respect to writes, which is why
you must use read-only access.  The texture cache is optimized for
noncontiguous access, since it has a shorter cache line than the
regular cache.

While \lstinline!RandomAccess! is valid for other execution spaces, 
currently no specific optimizations are performed. But in the future a view
allocated with the \lstinline!RandomAccess! attribute might for example
use a larger page size, and thus reducing page faults in the memory 
system.

@@

It's always better to let Kokkos control memory allocation, but
sometimes you don't have a choice.  You might have to work with an
application or an interface that returns a raw pointer, for example.
Kokkos lets you wrap raw pointers in an \emph{unmanaged View}.
``Unmanaged'' means that Kokkos does \emph{not} do reference counting
or automatic deallocation for those Views.  The following example
shows how to create an unmanaged View of host memory.  You may do this
for CUDA device memory too, or indeed for memory allocated in any
memory space, by specifying the View's execution or memory space
accordingly.
\begin{lstlisting}
// Sometimes other code gives you a raw pointer, ...
const size_t N0 = ...;
double* x_raw = malloc (N0 * sizeof (double));
{
  // ... but you want to access it with Kokkos.
  //
  // malloc() returns host memory, so we use the 
  // host memory space HostSpace.  Unmanaged
  // Views have no label, because labels work with
  // the reference counting system.
  View<double*, HostSpace, MemoryTraits<Unmanaged> > x_view (x_raw, N0);

@@
